---
title: "Cal_Merge: Script for Merging Independent Calorimetry Runs"
author: "Chelsea Faber"
format: 
   html:
     df-print: paged
editor: visual
---

## How to use this script:

###### Inputs:

-   Cohorts and associated run dates to be combined ([cohort1]{.underline}, [rundate1]{.underline}, [cohort2]{.underline}, [rundate2;]{.underline} type character)

-   binsize in minutes (type double)

-   csv and Rda specifying desired file type(s) for merged runs (type logical)

###### Outputs:

-   .Rda containing the trimmed and aligned runs

-   (optional) csv containing the trimmed and aligned runs

###### Known Bugs:

-   Rendering this .qmd will fail - there is some issue with the exporting to csv that prevents rendering in HTML. The code will still run properly within R Studio, however.

###### To Do:

1.  Add conditional loading of original binning if higher resolution than hourly is needed

```{r}
#| label: load-packages
#| warning: false
library(tidyverse)
library(magrittr)
library(lubridate)
library(here)
```

## Define inputs to script for analysis

```{r}
# First df to be merged
cohort1       <- "mon001" 
rundate1      <- "2021-10-18" 

# Second df to be merged
cohort2       <- "mon002" 
rundate2      <- "2021-11-03" 

# Resolution needed, in minutes. Default 60.
binsize       <- 60

# Specify how results should be exported. Default Rda only.
csv           <- FALSE
Rda           <- TRUE
```

## Load data frames

```{r}
run1         <- paste(rundate1,cohort1,"Clean.Rda", sep = "_")
run2         <- paste(rundate2,cohort2,"Clean.Rda", sep = "_")

load(here::here(run1))

# Only need df.hourly, so save this to df1 and df2 and remove all other data from workspace
df1          <- df.hourly
rm(list = setdiff(ls(), c("df1","cohort1","rundate1","cohort2","rundate2", "run2","Rda","csv")))

load(here::here(run2))
df2          <- df.hourly
rm(list = setdiff(ls(), c("df1","df2","cohort1","rundate1","cohort2","rundate2","Rda","csv")))
```

## Inspect data frames

Check that the column names are equivalent between the two runs to be combined. If a different macro version was used to analyze the raw .exp data, some differences may be present that will have to be manually resolved in the Cal_Clean.R script.

```{r}
if (!setequal(colnames(df1),colnames(df2))) {
  
  stop('Different column names between df1 and df2. Examine the data frames and rerun raw data through Cal_Clean.R if necessary.')

}
```

## Align and merge data frames

First, shift one data frames dates to align with the other. Note that this will shift the calendar dates, but not clock time.

```{r}
# Shift one data frame so that start dates align
start1 <- head(df1$DateTime,1)
start2 <- head(df2$DateTime,1)

if (start2 > start1) {
  ordered <- TRUE
  rec_lag <- as.integer(floor(start2 - start1))
  df2_shifted <- df2 %>% 
    mutate(DateTime = DateTime - days(rec_lag))
  df2_shifted <- df2_shifted[order(df2_shifted$DateTime),] %>%
    arrange(Animal)
  
} else {
  ordered <- FALSE
  rec_lag <- as.integer(floor(start1 - start2))
  df2_shifted <- df2 %>% 
    mutate(DateTime = DateTime + days(rec_lag))
  df2_shifted <- df2_shifted[order(df2_shifted$DateTime),] %>%
    arrange(Animal)
}
```

Next, find the indices of matching DateTimes. Create new data frames containing matched times only.

```{r}
# Find locations where DateTimes in df1 match values in df2 and vice versa
df1_match_idx <- match(df1$DateTime,df2_shifted$DateTime)
df2_match_idx <- match(df2_shifted$DateTime,df1$DateTime)

# Subset matching indices into new df_matched objects
df1_matched <- df1[!is.na(df1_match_idx),]
df2_matched <- df2_shifted[!is.na(df2_match_idx),]

# Recompute Time column to start from 0 in both tables.
start_time <- df1_matched$DateTime[1]
df1_matched %<>%
  mutate(Time = as.numeric(difftime(DateTime,start_time),units = "hours"))

df2_matched %<>%
  mutate(Time = as.numeric(difftime(DateTime,start_time),units = "hours"))
```

Combine data frames into one, stacking the tables with `bind_rows()`.

```{r}
# Final quality check before merging data frames
if (setequal(df1_matched$DateTime,df2_matched$DateTime)) {
  print("DateTimes are aligned, merging data frames into df_merge")
  df_merge <- bind_rows(df1_matched,df2_matched,.id = "Df_ID")
} else {
  print("Warning: DateTimes not aligned. Contact Chelsea for help!")
  paste("df1 has {length(unique(df1_matched$DateTime))} unique datetimes, while df2 has {length(unique(df2_matched$DateTime))} unique datetimes.")
}
```

## Export data

```{r}
#| warning: false
#| echo: false
filename <- paste("MERGED",rundate1,cohort1,rundate2,cohort2, sep = "_")

#/ warning: false
if (csv != FALSE) {
  print("Saving .csv with combined runs.")
  write.csv(df_merge, paste(filename, ".csv", sep = ""),row.names = FALSE)
} else {
  print("No .csv will be saved. Set csv equal to TRUE to change.")
}

if (Rda) {
  save(df_merge,file = paste(filename,".Rda", sep = ""))
  print("Saving .Rda with combined runs.")
} else {
  print("No .Rda will be saved. Set Rda equal to TRUE to change.")
}
```
