---
title: "Cal_Segment"
format: html
editor: visual
---

## About

**Note that this script (like the Cal_Clean script) has a default behavior of trimming the segmented data frames such that they each start and stop at 18:00 (ZT12).** This ensures that there are always an even number of L/D photoperiods, and that each always contains a full 12h, and guarantees subsequent statistical analysis will be more straightforward.

------------------------------------------------------------------------

## To do:

-   Improve speed/efficiency - vectorize code instead of for loops?

## Load required packages

```{r}
#| label: load-packages
#| warning: false
#| echo: false
library(plyr, include.only = 'mapvalues')
library(tidyverse)
library(magrittr)
library(lubridate)
library(here)
```

## Load data

```{r}
cohort          <- "CAL009"
rundate         <- "2022-07-11"
load(here::here(paste(rundate,cohort,"Clean.Rda",sep = "_")))
rm(list = setdiff(ls(), c("df","cohort","rundate", "df_code")))
```

## Prompt user for more information about reason for segmentation

```{r}
# Experimental parameters that could change
expvars         <- c("Treatment","Diet","Light","Temp")

# Prompt user to identify which, if any, parameters have changed
change          <- menu(c("Treatment","Diet","Light","Temp","none - segment only"), title = "Which experimental condition changed during the run?")

# If none change, exit loop
if (change == 5) {
  changevar     <- NA
  break
} else {
  changevar     <- expvars[[change]] # Index experimental name
  changetime    <- as.numeric(readline(prompt = paste("In units of hours from the start of the recording, when did the", changevar, "change?")))
  cages         <- menu(c("All","Custom"), title = "Which calorimetry cages were affected?")
  
  # Easiest case if all cages change the same way
  if (cages == 1) {
    cond1      <- readline(
      prompt = paste("What was the first condition for", changevar, "?"))
    cond2      <- readline(
      prompt = paste("What was the second condition for", changevar, "?"))
  } else {
    
    # Get more input on more complicated cases where there are more groups
    if (cages == 2) {
      
      n <- c(2,3,4)
      prompt <- "How many cage groupings exist for this intervention?"
      grouping   <- menu(c("2 groups","3 groups","4 groups","other"), title = prompt)
  
      # Prompt user to stop trying to automate for extreme cases
      if (grouping == 4) {
        print("Seek help. This is too custom for Chelsea to automate.")
      } else {
        
        # Create list with n groupings for user to enter cage #s
        ngroup <- n[grouping]
        groups <- vector(mode = "list", length = ngroup)
        for (i in 1:ngroup) {
          prompt <- paste("Please enter cages for group", as.character(i),"as a space-separated list.")
          temp <- strsplit(readline(prompt)," ")[[1]]
          temp <- as.integer(temp)
          groups[[i]] <- temp
          rm(temp)
        }
        
        # Add check to make sure all cages within unique(df$Cage) are in groups
        
        # If not, prompt user to reconcile discrepancy
        
        
      }
    }
  }
}


```

## Prompt user for segmentation times

```{# {r}
# span            <- last(unique(df$Time))
# print(paste("Recording spans", as.character(span), "hours."))
# n.segments      <- as.integer(readline(prompt = "Into how many segments should the run be split? "))
# 
# segtimes        <- data.frame(start = rep(0, each = n.segments), 
#                               stop = rep(0, each = n.segments))
# 
# for (i in 1 : n.segments) {
#   
#   segtimes[i,1] = as.integer(readline(prompt = 
#                                       paste("In units of hours from the start of the recording, where is the START of segment",as.character(i),"?")))
#   segtimes[i,2] = as.integer(readline(prompt = 
#                                       paste("In units of hours from the start of the recording, where is the STOP of segment",as.character(i),"?")))
#   
# }
```

## Filter df into segments

```{r}
df.segments     <- vector(mode = "list", length = n.segments)
for (i in 1 : n.segments) {
  
  start <- segtimes[i,1]
  stop <- segtimes[i,2]
  df.segments[[i]] = df %>% filter(Time >= start & Time <= stop)
  
}
#rm(df)
```

## Trim new segments to start/stop at 18:00

```{r}
for (i in 1 : n.segments) {
  
  tempdf <- df.segments[[i]] 
  
  # Find first lights out (18:00) for each animal
  tempdf %<>% group_by(Animal)
  idx.start        <- which(hour(tempdf$DateTime) == 18)[1]
  time.start       <- as.POSIXct(tempdf$DateTime[idx.start])
  
  # Filter
  tempdf %<>% filter(DateTime >= time.start)
  
  # Find last lights out
  idx.end          <- tail(which(hour(tempdf$DateTime) == 17), n=1)
  if (idx.end != nrow(tempdf)) {
    idx.end <- idx.end+1
  }
  time.end         <- as.POSIXct(tempdf$DateTime[idx.end])

  # Filter again
  tempdf %<>% filter(DateTime < time.end) %>% ungroup() %>% mutate(Segment = as.integer(i))
  
  df.segments[[i]] <- tempdf
  rm(tempdf)
}
```

## Restart the clock & recompute cumulative variables

```{r}
for (i in 1 : n.segments) {
  
    tempdf <- df.segments[[i]]
    tempdf %<>% group_by(Animal) %>%
      mutate(
        ExpDay = as.integer(ceiling(difftime(DateTime,time.start, 
                                             units = "days"))),
        Time = as.numeric(difftime(DateTime,time.start), 
                          units = "hours"),.after = DateTime) %>% 
      mutate(FoodIn.cum = cumsum(FoodIn.g)) %>%
      mutate(FoodIn.cum.kcal = case_when(
        Diet == "NCD" ~ FoodIn.g * 3.35,
        Diet == "HFD" ~ FoodIn.g * 5.47)) %>% 
      mutate(WaterIn.cum = cumsum(WaterIn.g)) %>% 
      mutate(EE.cum = cumsum(EE.kcal.bin)) %>%
      mutate(EB.cum = cumsum(EBalance)) %>%
      mutate(AllMeters.cum = cumsum(AllMeters)) %>% 
      ungroup()
  
}
```
